---
title: "LoanRiskAnalysis"
author: "Liang Tan"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Read data
```{r tidy=TRUE}
loan <- read.csv('loan.csv',  stringsAsFactors = FALSE) 
loanT <- loan
head(loan)
```
Check dimension
```{r tidy=TRUE}
print(dim(loan))
```
Check data format
```{r tidy=TRUE}
str(loan)
```
       
Calculate the number of na values for each column.    
```{r tidy=TRUE}
#seperate columns with over 80% NA values
num.NA <- sort(sapply(loan, function(x) sum(is.na(x))), decreasing = TRUE) 
remain.col = names(num.NA)[(num.NA < 0.8 * dim(loan)[1])]
delete.col = names(num.NA)[(num.NA >= 0.8 * dim(loan)[1])]
delete.col
```

# EDA_part_1 (What factor will influence the interest rate?)   

```{r tidy=TRUE}
{plot(density((loan$int_rate)), main='Density plot of interest rate', xlab = 'Interest rate')
abline(v = mean(loan$int_rate), lty=2, col = 2)
abline(v = median(loan$int_rate), lty=2, col=3)
abline(v = quantile(loan$int_rate, 0.2), lty=2, col=4)
abline(v = quantile(loan$int_rate, 0.8), lty=2, col=6)
legend("topright", c('mean', 'median', '20% quantile', '80% quantile'), col = c(2,3,4,6), lty = 2, title = 'Interest rate statistics')}
```

The distribution is a bit right skew. In the future, if we want to build model with interest rate then it is better to use square root to adjust the skewness. Next I want to explore the correlation between interest rate with other numeric variables. However, we know some features are with high number of NA values. So I want to remove those features with over 80% NA values temperarily.     

```{r tidy=TRUE}
library(corrplot)
loan <- loan[,remain.col]
#select numerical features' name
num.feature <- names(loan[, sapply(loan, is.numeric)])    
#select char features' name
char.feature <- names(loan[, sapply(loan, is.character)]) 
#calculate the correlationship between int_rate and other numerical features
correlation <- cor(loan$int_rate, loan[, num.feature], use = 'pairwise.complete.obs') 
#sort the value
correlation <- correlation[, order(correlation[1,], decreasing = TRUE)] 
#remove correlation with itself and correlation with police_code
correlation <- correlation[2:33] 
```

```{r tidy=TRUE}
{barplot(correlation, main = 'Correlation between int_rate with other numerical features', ylab = 'Correlation', las=2,
        cex.names=0.6, col=ifelse(correlation>0.15,"red","blue"))
legend("topright", leg = c('top five correlated feature', 'other numerical feature'), fill = c("red", "blue"))}
```
      
Based on correlation calculation, the top five predictive numerical features for int_rate are "total_rec_int", "revol_util", "inq_last_6mths", "total_pymnt_inv" and "total_pymnt". Next step is to explore the top five influencial category features.  

```{r tidy=TRUE}
library(ggplot2)
ggplot(data=loan, aes(term,int_rate,fill=term))+geom_boxplot(outlier.color = "blue")+labs(title="Box plot of interest rate against term") + theme(axis.text.x = element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank(), plot.title = element_text(hjust = 0.5))
```
```{r tidy=TRUE}
t.test(subset(loan, term == ' 36 months')$int_rate, subset(loan, term == ' 60 months')$int_rate, conf.level = 0.95, mu = 0, alternative = "two.sided", paired = FALSE, var.equal = FALSE)
```
       
There is a significant difference between different term. Therefore, term can be used a predictor for interest rate.   

```{r tidy=TRUE}
ggplot(data=loan, aes(grade,int_rate,fill=grade))+geom_boxplot(outlier.color = "blue")+labs(title="Box plot of Interest rate against grade") + theme(axis.text.x = element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank(), plot.title = element_text(hjust = 0.5))
```
      
There is a clear linear trend between interest rate and grade. Therefore, grade can be used a predictor for interest rate. However, based on the description of grade. It is assigned by the Lending Club. Therefore, I probably don't have this feature in advance. If a client is a return user and I can definitely use this information.     

```{r tidy=TRUE}
ggplot(data=loan, aes(verification_status,int_rate,fill=verification_status))+geom_boxplot(outlier.color = "blue")+labs(title="Box plot of interest rate against verification_status") + theme(axis.text.x = element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank(), plot.title = element_text(hjust = 0.5))
```
```{r tidy=TRUE}
t.test(subset(loan, verification_status == 'Verified')$int_rate, subset(loan, verification_status == 'Source Verified')$int_rate, conf.level = 0.95, mu = 0, alternative = "two.sided", paired = FALSE, var.equal = FALSE)
t.test(subset(loan, verification_status == 'Not Verified')$int_rate, subset(loan, verification_status == 'Source Verified')$int_rate, conf.level = 0.95, mu = 0, alternative = "two.sided", paired = FALSE, var.equal = FALSE)
t.test(subset(loan, verification_status == 'Not Verified')$int_rate, subset(loan, verification_status == 'Verified')$int_rate, conf.level = 0.95, mu = 0, alternative = "two.sided", paired = FALSE, var.equal = FALSE)
```
         
There is a significant difference between different verification_status Therefore, verification_status can be used a predictor for interest rate.       

```{r tidy=TRUE}
ggplot(data=loan, aes(pymnt_plan,int_rate,fill=pymnt_plan))+geom_boxplot(outlier.color = "blue")+labs(title="Box plot of interest rate against pymnt_plan") + theme(axis.text.x = element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank(), plot.title = element_text(hjust = 0.5))
```
```{r tidy=TRUE}
t.test(subset(loan, pymnt_plan == 'n')$int_rate, subset(loan, pymnt_plan == 'y')$int_rate, conf.level = 0.95, mu = 0, alternative = "two.sided", paired = FALSE, var.equal = FALSE)
```
          
There is a significant difference between different pymnt_plan. Therefore, pymnt_plan. can be used a predictor for interest rate.        

```{r tidy=TRUE}
ggplot(data=loan, aes(application_type,int_rate,fill=application_type))+geom_boxplot(outlier.color = "blue")+labs(title="Box plot of interest rate against application_type") + theme(axis.text.x = element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank(), plot.title = element_text(hjust = 0.5))
```
```{r tidy=TRUE}
t.test(subset(loan, application_type == 'INDIVIDUAL')$int_rate, subset(loan, application_type == 'JOINT')$int_rate, conf.level = 0.95, mu = 0, alternative = "two.sided", paired = FALSE, var.equal = FALSE)
```
           
There is a significant difference between different application_type. Therefore, application_type can be used a predictor for interest rate. In conclustion, these five category variables are influenciable: "term", "grade", "verification_status", "pymnt_plan" and "application_type". Besides I am also curious about how interest rate vary with space and time.       

```{r tidy=TRUE}
library(dplyr)
suppressPackageStartupMessages(library(maps))
loan$region <- loan$addr_state
loan$region <- as.factor(loan$region)
levels(loan$region)<- c("alaska", "alabama","arkansas", "arizona", "california","colorado","connecticut","district of columbia","delaware","florida","georgia","hawaii","iowa","idaho","illinois","indiana","kansas","kentucky","louisiana","massachusetts","maryland","maine","michigan","minnesota","missouri","mississippi","montana","north carolina","north dakota","nebraska","new hampshire","new jersey","new mexico","nevada","new york","ohio","oklahoma","oregon","pennsylvania","rhode island","south carolina","south dakota","tennessee","texas","utah","virginia","vermont","washington","wisconsin","west virginia","wyoming")

all_states <- map_data("state")
state_by_rate <-loan %>% group_by(region) %>%
                summarise(value = mean(int_rate, na.rm=TRUE))
state_by_rate$region <- as.character(state_by_rate$region)

Total <- merge(all_states, state_by_rate, by="region")

p <- ggplot()
p <- p + geom_polygon(data=Total, aes(x=long, y=lat, group = group, fill=Total$value),colour="white"
      ) + scale_fill_continuous(low = "skyblue", high = "darkblue", guide="colorbar")
P1 <- p + theme_bw()  + labs(fill = "Interest rate" 
                            ,title = "Heat map of interest rate in all states", x="", y="")
P1 + scale_y_continuous(breaks=c()) + scale_x_continuous(breaks=c()) + theme(panel.border =  element_blank(), plot.title = element_text(hjust = 0.5))
```
         
# EDA_part_2 (What are the distribution of loan status?)         

```{r tidy=TRUE}
library(DescTools)
Desc(loan$loan_status, plotit = TRUE, main = "Loan status")
```
           
See how does loan_status affact interest rate.    

```{r tidy=TRUE}
ggplot(data=loan,aes(loan_status,int_rate), las=2) + geom_boxplot(aes(fill=loan_status))+labs(list(title = "Interest rate by loan_status",x = "Loan_status",y = "Interest rate"))  +theme(axis.text.x = element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank(), plot.title = element_text(hjust = 0.5))
```            
           
The majority status is "Current". The finished loan can be grouped into "Fully Paid" and "Charged off" or "Late payment".       

# EDA_part_3 (What are the purpose of applying a loan with Lending Club ?)       

```{r tidy=TRUE}
Desc(loan$purpose, main = "Loan purposes", plotit = TRUE)
```
    
I am also curious about how interest rate vary with purpose.      

```{r tidy=TRUE}
ggplot(data=loan,aes(purpose,int_rate), las=2) + geom_boxplot(aes(fill=purpose))+labs(list(title = "Interest rate by purpose",x = "purpose of Loan",y = "Interest rate"))  +theme(axis.text.x = element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank(), plot.title = element_text(hjust = 0.5))
```


